

# Communicable, Maternal, Neonatal, and Nutritional Disease Burden (CMNN)
#### Adrián Calvo-Valderrama




```{r}
# Load necessary packages
pacman::p_load(tidyverse, knitr, here, janitor)

# Import the communicable diseases data
data_cmnn <- read_csv(here("data/burden-of-disease-cmnn.csv"))
```

```{r}
data_cmnn_1 <- data_cmnn %>% 
  filter(Entity == "Liberia" | Entity == "United States" | Entity ==  "Spain") %>% 
  rename(dalys = `DALYs (Disability-Adjusted Life Years) - Communicable, maternal, neonatal, and nutritional diseases - Sex: Both - Age: Age-standardized (Rate)`) %>% 
  clean_names()
```

## Table of Estimates for CMNN Burden Over Time

```{r}
# Here render a table for the DALY burden over time for the three countries.
# You should pivot the data to show each country in a separate column.
# Use kable() from the knitr package to render the table.
library(knitr)

table1 <- data_cmnn_1 %>% 
  select(-code) %>% 
  pivot_wider(names_from = entity,
            values_from = dalys)

kable(table1)
```

## Plot Showing Trends in CMNN Burden Over Time

```{r}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.

ggplot(data_cmnn_1,
       mapping = 
         aes(x = year,
             y= dalys,
             colour = entity)) +
  
  geom_line() +
    labs(x = "Year",
         y = "DALYs",
       title = "Communicable, Maternal, Neonatal, and Nutritional Disease Burden (CMNN)",
       subtitle = "Liberia, United States, and Spain, 1990-2019",
       colour = "Country")



```


## Plot Showing Cumulative Trends in CMNN Burden Over Time

```{r}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.

ggplot(data_cmnn_1,
       mapping = 
         aes(x = year,
             y= cumsum(dalys),
             colour = entity)) +
  
  geom_line() +
    labs(x = "Year",
         y = "DALYs",
       title = "Cumulative CMNN Disease Burden",
       subtitle = "Liberia, United States, and Spain, 1990-2019",
       colour = "Country")



```



## Summary of CMNN Burden Findings

Shown above is the burden of Communicable, Maternal, Neonatal, and Nutritional Disease Burden (CMNN) for Liberia, Spain and the United States between 1990 and 2019.The graph above shows that whilst the United States and Spain display a pretty trend over this time period. Meanwhile, whilst Liberia's CMNN Disease Burden numbers are significantly higher than the other two countries, it shows a strong decrease over the same period, with a temporary uptick around 2013.