# Non-Communicable Disease Burden (NCD)

```{r}
# Load necessary packages
pacman::p_load(tidyverse, knitr, here, janitor)

# Import the non-communicable diseases data
data_ncd <- read_csv(here("data/burden-of-disease-ncd.csv"))
```

```{r}
# Filter the dataset to include only the three countries chosen by your group.
filtered_data <- data_ncd %>%
  filter(Entity == "Liberia" | Entity == "Spain" | Entity == "United States") %>% 
  rename(dalys = 4) %>% 
  clean_names()

```


## Table of Estimates for NCD Burden Over Time

```{r}
# Here render a table for the DALY burden over time for the three countries. 
# You should pivot the data to show each country in a separate column.
# Use kable() from the knitr package to render the table.

daly_table <- filtered_data %>% 
  select(-code) %>% 
pivot_wider(names_from = entity, values_from = dalys)

kable(daly_table)

```

## Plot Showing Trends in NCD Burden Over Time 

```{r}
# Here, plot a line graph showing the trend of DALY burden for the chosen countries over time. You can use ggplot2 to create this plot. Each country's line should be a different color.
daly_plot <- filtered_data %>% 
  ggplot(aes(x = year, y = dalys, color = entity)) +
  geom_line() +
  labs(x = "Year", 
       y = "DALYs", 
       color = "Country", 
       #title = "Non-Communicable Disease (NCD) Burden",
       #subtitle = "Liberia, United States, and Spain, 1990-2019"
       )
daly_plot
```

## Summary of NCD Burden Findings

Shown above is the burden of non-communicable disease (NCD) in Liberia, Spain and the United States from 1990 to 2020. Spain is unique, as it shows a fairly linear downward trend in NCD burden. The burden of disease in Liberia is also generally trending downward, though there is a notable spike in the mid 2010's. From 1990-2010 the downward trend of NCD burden in the United States has been less dramatic than both Liberia and Spain. Notably, the United States has been experiencing an increase in NCD DALYs since the mid 2010's. While a growing elderly population and increased prevalence of age-related disease might explain the very slight leveling of the curve in Spain, it is unlikely that the increase in NCD burden in the United States can be explained by this alone. It is plausible that a major contributing factor is widespread issues with healthcare access, especially preventative and routine care.


